Parallelization of Fast $\ell_1$-Minimization for Face Recognition

### 2011-06-05 main.tex
Merge changes from Allen

\begin{abstract} Recently a family of promising face recognition algorithms
based on sparse representation and $\ell_1$-minimization ($\ell_1$-min) have been
developed.  These algorithms have not yet seen commercial application, 
largely due to higher computational cost compared to other traditional
algorithms. This paper studies techniques for leveraging the 
massive parallelism available in GPU and CPU hardware to accelerate
$\ell_1$-min based on augmented Lagrangian method (ALM) solvers. 
Due to the low arithmetic intensity of ALM solvers, both
architectures are bandwidth limited.  
with large problems running faster on the GPU, and small problems running faster
on the CPU.  
On both architectures, the proposed implementations dramatically out-perform
naive library-based implementations. 
At the problem sizes most important for $\ell_1$-min based
face recognition, both architectures achieve comparable performance.
The source code of the
algorithms will be made available for peer evaluation.  \end{abstract}

\section{Introduction} 
$\ell_1$-minimization ($\ell_1$-min) has received much attention in recent
years due to important applications in compressive sensing
\cite{BrucksteinA2007} and sparse representation \cite{WrightJ2010-PIEEE}.  
The problem refers to finding the minimum $\ell_1$-norm solution to an
underdetermined linear system $\bb=A\xx$:
\begin{equation}
\min \|\xx\|_1\quad \mbox{ subj. to }\quad \bb = A\xx.
\label{eq:l1min}
\end{equation}
It is now well established that, under certain conditions
\cite{CandesE2005-IT_1,DonohoD2004}, the minimum $\ell_1$-norm solution is also
the \emph{sparsest} solution to the system \eqref{eq:l1min}.

Among its many applications, $\ell_1$-min has been recently used to reformulate
image-based face recognition as a sparse representation problem
\cite{WrightJ2009-PAMI}.  If we stack the pixels of the training images of $K$ subject
classes into the columns of matrices $(A_1\in\Re^{d\times n_1}, \cdots, A_K\in\Re^{d\times n_K})$, combine
the matrices into a larger matrix $A = [A_1, \cdots, A_K]\in\Re^{d\times n}$, and arrange the pixels of a new
query image into a vector $\bb\in\Re^d$, \emph{sparsity-based
classification} (SBC) solves the following minimization problem:
\begin{equation}
\min_{\xx, \ee} \| \xx \|_1 + \|\ee\|_1 \quad \subj \quad \bb = A \xx + \ee.
\label{eq:l1min_denoise}
\end{equation}
If the sparsest solutions for $\x$ and $\e$ are recovered, $\ee$ provides a
means to compensate for pixels that are corrupted due to occlusion of some part of the query
image, and the dominant nonzero coefficients of $\xx$ reveal the membership of
$\bb$ based on the training image labels associated with $A$.  Since $A$
contains many images per user taken under different illuminations, 
$A_i\x_i$ acts as a linear illumination model for the test image $\bb$.

@@

While solving \eqref{eq:l1min_denoise} can achieve exceptional
recognition accuracy on public databases, its success also relies on the assumption
that the query image is well aligned with the training images. In
\cite{WagnerA2009-CVPR}, the framework of SBC was extended to align
a query image to each subject class individually. The face alignment algorithm iteratively solves
the following minimization problem:
\begin{equation}
\hat{\tau}_i = \arg\min_{\xx, \ee, \tau_i} \|\ee\|_1\quad \mbox{subj. to}\quad \tilde\bb\circ\tau_i = A_i\xx + \ee,
\label{eq:l1min_alignment}
\end{equation}
where $\tau_i\in T$ is in a finite-dimensional group $T$ of transformations
(e.g., affine or homography) acting on the image domain.  The sequence
$\tau_i$ converges to a transformation that aligns the test image $\tilde\bb$
with the training images from the $i$-th class $A_i$. In other words, the algorithm 
extends Lucas-Kanade iterative alignment \cite{LucasB1981} to use the
$\ell_1$-norm as a robust error function, while simultaneously estimating
the coefficients of the linear illumination model $A_i\x$.

Numerically, the original solution \cite{WagnerA2009-CVPR} was implemented based 
on standard interior-point methods, and its speed has been since improved by ALM in a more recent paper
\cite{WagnerA2011-PAMI}. However, due the high per-class cost
of the alignment step, the recognition system still fails to achieve real-time
recognition against datasets of hundreds or thousands of subjects.  In this paper,
in addition to accelerating the generic
$\ell_1$-min objectives \eqref{eq:l1min} and \eqref{eq:l1min_denoise}, we 
also discuss how to efficiently accelerate the iterative face alignment step
\eqref{eq:l1min_alignment} on multi-core CPUs and GPUs.

@@

Finally, we present extensive benchmark to compare the performance of the
generic ALM $\ell_1$-min algorithm and the corresponding face recognition
routines on massively parallel CPUs and GPUs.  To this end, a face
recognition system for security and access control applications has been implemented, 
and is diagrammed in Figure \ref{fig:pipeline}. This pipeline follows the work of \cite{WagnerA2011-PAMI}, but
the fast implementations presented in this paper are novel.
Since the parallelization of any algorithm
is highly dependent on the target hardware, we will give a brief overview of
the architecture of our test system in detail in Section
\ref{sec:concurrency}.
Finally, note that 
while this paper focuses on the parallelism available in a single (shared memory) machine,
applications targeting thousands of subjects or more, will likely require computation be
distributed over a cluster or network of computers, and this is not addressed here.
The source code of our implementations, which target NVIDIA Fermi GPUs
as well as Intel x86-64 CPUs, will be made available for peer evaluation.
\begin{figure}
\centering
{\tiny \input{figures/pipeline_simplified.tex}}
\caption{The face recognition pipeline.}
\label{fig:pipeline}
\end{figure}

\subsection{Literature Review} 
Traditionally, $\ell_1$-min (a.k.a.
basis pursuit (BP)) has been formulated as a linear program
\cite{ChenS2001-SIAM}. Several variations of the solution are also well known
in optimization, including a noisy approximation via quadratic programming
called the LASSO \cite{TibshiraniR1996} and truncated Newton interior-point
method (TNIPM) \cite{KimS2007}.

@@

\section{Augmented Lagrangian Method}
\label{sec:ALM}

In this section, we briefly describe the ALM algorithm for $\ell_1$-min
\eqref{eq:l1min} \cite{YangA2010-ICIP} and analyze its complexity. Lagrange
multipliers have been frequently used in convex programming to eliminate
equality constraints via adding a penalty term to the cost function for
infeasible points. ALM methods differ from other penalty-based approaches by
simultaneously estimating the optimal solution and Lagrange multipliers in an
iterative fashion.  For $\ell_1$-min \eqref{eq:l1min}, the augmented Lagrange
function is defined as: 
\begin{equation} L_\mu(\xx,\yy) = \|\xx\|_1 +
\yy^T(\bb - A\xx) + \frac{\mu}{2} \| \bb-A\xx \|_2^2, 
\end{equation}
where $\mu > 0$ is a constant that penalizes infeasibility and $\yy$ is a
vector of Lagrange multipliers.

@@

 \begin{algorithm}[h]
\caption{Augmented Lagrangian Method (ALM)}
{\bf INPUT:} $\bb \in \Re^m$, $A=[A_1,\cdots, A_K] \in \Re^{m \times n}$, $\tau\leftarrow \max\mbox{eig}(A^TA)$, and constant $\rho>1$.
\begin{algorithmic}[1]
\WHILE{not converged ($k = 1,2,\ldots$)} 
\STATE $t_1 \leftarrow 1$, $\zz_1 \leftarrow \xx_k$, $\uu_1 \leftarrow \xx_k$ 
\WHILE{not converged ($l = 1,2,\ldots$)} 
\STATE $\uu_{l+1}  \leftarrow \mbox{shrink}(\zz_l - \frac{1}{\tau}A^T(A\zz_l - \bb - \frac{1}{\mu_k}\yy_k), \frac{1}{\mu_k\tau})$
\STATE $t_{l+1} \leftarrow \frac{1}{2}( 1 + \sqrt{1+4t_l^2})$
\STATE $\zz_{l+1} \leftarrow \uu_{l+1}+ \frac{t_l - 1}{t_{l+1}}(\uu_{l+1} - \uu_l)$ 
\ENDWHILE 
\STATE $\xx_{k+1} \leftarrow \uu_{l+1}$ 
\STATE $\yy_{k+1} \leftarrow \yy_k + \mu_k (\bb - A\xx_{k+1})$ 
\STATE $\mu_{k+1} \leftarrow \rho\cdot\mu_k$ 
\ENDWHILE 
\end{algorithmic}

@@

\section{Hardware Concurrency} \label{sec:concurrency}

In this section we discuss the levels of concurrency available in the hardware
architectures considered in this paper, as well as other aspects of the
hardware that are important for performance.  In particular, since caches
(regions of on-chip memory) are often orders of magnitude faster than off-chip
memory, their sizes and speeds have a dramatic effect on performance.  We give a
brief overview of the caches that are available in our target architectures,
and defer discussion of their performance effects to Sections
\ref{sec:alignment_implementation_cpu} and
\ref{sec:alignment_implementation_gpu}.

Our discussion and experiments will address the most common hardware
configuration for engineering workstations: a motherboard with two quad-core
processors, and a PCI card with a single high-end GPU.  
Recognition involving more than a few hundred subjects with contemporary hardware
would require a server(or cluster) configuration with an expandable 
number of processors.  We will not address efficient parallelization for 
these systems, which may have additional challenges associated with
their non-shared memory model.\footnote{Non-shared memory configurations
include CPU blade servers, multi-GPU servers, and clusters}.
Unless otherwise specified, all implementations utilize single precision
floating point datatypes.  
% WORKING HERE

\subsection{CPU Hardware Concurrency} 
\label{sec:CPU-concurrency}
The main defining characteristics of contemporary multi-core CPU architectures
are that they have two levels of concurrency, relatively large amounts of
cache, and relatively high clock speeds.

The baseline architecture for our experiments is a Linux workstation with two
quad-core Intel Nehalem E5530 processors clocked at 2.4 GHz.  Each processor
has its own memory interface, and is directly interfaced to half of the RAM
installed in the machine.  The amount of RAM installed exceeds the amount used
by the algorithms, and is not an important performance consideration.

\begin{figure}
\centering
\subfigure[The larger algorithm data structures]
{
	\includegraphics[scale=1.2]{figures/arrays}
	\label{fig:arrays}
}
\subfigure[The caches on a E5530 CPU]
{
	\includegraphics[scale=1.2]{figures/cpu_caches}
	\label{fig:cpu_caches}
}
\subfigure[The caches on a GTX480 GPU]
{
	\includegraphics[scale=1.2]{figures/gpu_caches}
	\label{fig:gpu_caches}
}
\label{fig:caches}
\caption{A visual comparison of the algorithm working set to the CPU and GPU
caches.  Note: Although the aspect ratio could not be preserved, all arrays and
caches are drawn so that the area is proportional to the size of the data. }
\end{figure}

@@

For floating-point instructions, each core also has a vector processing unit
(SSE) capable of performing the same arithmetic operation on four single-precision 
floating point values simultaneously.
There are thus two important levels of concurrency that need to be exploited to
efficiently use a modern CPU: {\em core-level} concurrency and {\em sse-level}
concurrency.

\subsection{GPU Hardware Concurrency}
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
 cache, and relatively low clock speeds.  Whereas most of the transistors on
a typical CPU are dedicated either to cache or to hardware that enables higher
clock speeds (such as branch prediction, out-of-order instruction execution, etc...),
most of the transistors on a GPU are dedicated to arithmetic logic units.
For our GPU implementations we target NVIDIA Fermi GPU architecture (\eg, the GTX 480
GPU used in this paper).  An explanation of the GPU programming model (CUDA) at a useful level of completeness would take more space than is available here, so we will instead frame our discussion in terms of 
hardware capabilities. \footnote{A discussion of the CUDA programming model can be found
in the NVIDIA C for CUDA Programming Guide}
CUDA programmable GPUs are comprised of several
\emph{streaming multiprocessors} (SMP), each of which is roughly analogous to a CPU core.
For Fermi architecture GPU's, there are up to 16 SMP's, and each SMP is capable of executing up to 64 single precision 
floating point operations
concurrently. \footnote{In CUDA terms, the warp width is 32, and the floating pipeline can
issue up to two warps simultaneously}

@@

While the programming model for
the SIMD units of each SMP is somewhat more flexible than on the CPU,
leveraging the flexibility typically comes at the cost of reduced concurrency
.\footnote{In CUDA terms, code branches that cause warp divergence usually
result in serialization of the different groups of threads}  Therefore, for
the purposes of comparing hardware architectures, the SIMD units on the GPU are
roughly analogous to the CPU SIMD units. Thus, in summary, the GPU hardware 
provides two levels of concurrency: {\em SMP-level}
concurrency and {\em thread-level} concurrency.  Note that the GPU provides more
concurrency than a CPU at both levels (14 SMP's vs. 8 cores) and (64 wide vs. 4 wide SIMD
units).

\section{Parallelism in the Face Alignment Stage}
\label{sec:alignment}

@@

Face alignment \eqref{eq:l1min_alignment} estimates an image transformation
$\tau$ that rectifies the query image $\tilde{\bb}$ with possible pose
variation w.r.t. each training class $A_i$, which leads to the minimal sparsity
in error $\ee$ after the alignment. Note that directly solving
\eqref{eq:l1min_alignment} is impossible since it is a non-convex problem
and may exhibit local minima.
However, given a good initial guess of the
transformation $\tau$ (\eg, provided by an accurate face detector), the optimal
solution for $\tau$ can be sought iteratively by linearization at each $j$th
step:
\begin{equation}
\min_{\xx, \ee, \Delta \tau_j}\|\ee\|_1\quad \subj\quad \tilde\bb\circ\tau_j +  J_j\Delta \tau_j = A_i\xx + \ee,
\end{equation}
where $J_j = \nabla_{\tau_j}(\tilde\bb\circ\tau_j)$ is the Jacobian, and
$\Delta \tau$ is a step update to $\tau$. Denote $\bb_j =
\tilde\bb\circ\tau_j$, $B_j = [A_i, -J_j]$ and $\ww^T = [\xx^T, \Delta
\tau^T]$, then the update $\Delta \tau$ can be computed by solving the
following linear program:
\begin{equation}
\min_{\ww, \ee}\|\ee\|_1\quad \subj\quad \bb_j = B_j\ww + \ee.
\label{eq:alignment-linearization}
\end{equation}
%If the warping $\tau$ is constrained to be a similarity transform,
%it can be paramaterized in such a way that the computation of $\frac{\partial \bb}{\partial \tau_p}$
%reduces to several linear vector operations involving pixel coordinates $(u,v)$, $\tau \in \Re^4$,
%and the image gradients $\f_x, \f_y \in \Re^M$, the latter of which only has to be computed
%once per test image.  
The per-class alignment algorithm via ALM is summarized in Algorithm
\ref{alg:iterative_alignment}. 
%In terms of computational complexity, one significant
%term is the computation of the matrix pseudo-inverse $B_j^\dagger$, whose
%computational cost is bounded by $O(n_i^2m + n_i^3)$. However, for per-class face
%alignment, since both $B_j$ and $J_{j}$ are very tall matrices (\ie, $m\gg n_i$
%for all $i=1, \cdots, K$ classes), we find that the computation cost is still
%dominated by the frequently computed matrix-vector multiplications in the
%inner loop.
\begin{algorithm}[ht!]
\caption{\bf (Face Alignment via ALM)} \label{alg:iterative_alignment}
{\bf Input:} $\bb$, $A_i$, $\x_0 = \mathbf{0}$, $\tau_0$, and $J_0$.
\begin{algorithmic}[1]
\WHILE{not converged ($j = 1,2,\ldots$)}
\STATE Update $\bb_j \leftarrow \frac{\bb\circ \tau_{j-1}}{\|\bb\circ \tau_{j-1}\|}$; $B_j= [A_i, -J_{j-1}]$ and corresponding $(B_j^\dagger)^T$
\STATE Initialize $\ww_0 = \mathbf 0$, $\yy_0 = \mathbf 0$
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\uu_0\leftarrow \ww_{k-1}$; $\zz_0\leftarrow \ee_{k-1}$
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\zz_l \leftarrow \mbox{shrink}\left(\bb_j - B_j\uu_{l-1} + \frac{\yy_{k}}{\mu_{k-1}}, \frac{1}{\mu_{k-1}}\right)$
\STATE $\uu_l \leftarrow B_j^\dagger \left(\bb_j - \zz_{l} + \frac{\yy_{k-1}}{\mu_{k-1}} \right) $
\ENDWHILE
\STATE $\ww_k \leftarrow \uu_l$; $\ee_k \leftarrow \zz_l$
\STATE $\yy_{k} \leftarrow \yy_{k-1} + \mu_{k-1} (\bb_j - B_j\ww_{k} - \e_{k})$
\STATE $\mu_{k} \leftarrow \rho\mu_{k-1}$
\ENDWHILE
\STATE Update $\ee_j$, $\tau_j$, and $J_j$
\ENDWHILE
\end{algorithmic}
{\bf Output:} $\tau_i^*\leftarrow \tau_j, \e_i^*\leftarrow \ee_j$
\end{algorithm}

@@

\subsection{CPU Implementation} 
\label{sec:alignment_implementation_cpu}

@@

\subsection{Recognition Stage Implementation} Compared to problem-level
parallelism, the exclusively pixel-level parallelism in Algorithm \ref{alg:alm_rec} is
relatively straightforward to exploit:  
On the CPU, most of the operations map
well onto MKL BLAS calls, and operations that do not can be easily
parallelized using OpenMP and auto-vectorization.

>>>>

\begin{abstract} Recently a family of promising face recognition algorithms
based on sparse representation and $\ell_1$-minimization ($\ell_1$-min) have been
developed.  These algorithms have not yet seen commercial application, 
largely due to higher computational cost compared to other traditional
algorithms. This paper studies techniques for leveraging the 
massive parallelism available in GPU and CPU hardware to accelerate
$\ell_1$-min based on augmented Lagrangian method (ALM) solvers. 
Due to the low arithmetic intensity of ALM solvers, implementations on both
architectures are primarily limited by memory bandwidth,  
with large problems running faster on the GPU and small problems running faster
on the CPU.  
On both architectures, the proposed implementations significantly outperform
naive library-based implementations. 
At the problem sizes most important for $\ell_1$-min based
face recognition, both architectures achieve comparable performance.
The source code of the
algorithms will be made available for peer evaluation.  \end{abstract}
\vspace{-0.1in}

\section{Introduction} 
\vspace{-0.05in}
$\ell_1$-minimization ($\ell_1$-min) has received much attention in recent
years due to important applications in compressive sensing
\cite{BrucksteinA2007} and sparse representation \cite{WrightJ2010-PIEEE}.  
The problem refers to finding the minimum $\ell_1$-norm solution to an
underdetermined linear system $\bb=A\xx$:
\begin{equation}
\min \|\xx\|_1\quad \mbox{ subj. to }\quad \bb = A\xx.
\label{eq:l1min}
\end{equation}
It is now well established that, under certain conditions
\cite{CandesE2005-IT_1,DonohoD2004}, the minimum $\ell_1$-norm solution is also
the \emph{sparsest} solution to the system \eqref{eq:l1min}.

Among its many applications, $\ell_1$-min has been recently used to reformulate
image-based face recognition as a sparse representation problem
\cite{WrightJ2009-PAMI}.  If we stack the pixels of the training images of $K$ subject
classes into the columns of matrices $(A_1\in\Re^{d\times n_1}, \cdots, A_K\in\Re^{d\times n_K})$, combine
the matrices into a larger matrix $A = [A_1, \cdots, A_K]\in\Re^{d\times n}$, and arrange the pixels of a new
query image into a vector $\bb\in\Re^d$, \emph{sparsity-based
classification} (SBC) solves the following minimization problem:
\begin{equation}
\min_{\xx, \ee} \| \xx \|_1 + \|\ee\|_1 \quad \subj \quad \bb = A \xx + \ee.
\label{eq:l1min_denoise}
\end{equation}
If the sparsest solutions for $\x$ and $\e$ are recovered, $\ee$ provides a
means to compensate for pixels that are corrupted due to occlusion of some part of the query
image, and the dominant nonzero coefficients of $\xx$ reveal the membership of
$\bb$ based on the training image labels associated with $A$. 
%Since $A$ contains many images per user taken under different illuminations, 
%$A_i\x_i$ acts as a linear illumination model for the test image $\bb$.

@@

In addition, the accuracy of the solution given by \eqref{eq:l1min_denoise} relies on the assumption
that the query image is well aligned with the training images.
In \cite{WagnerA2009-CVPR}, SBC was extended to align
a query image to each subject class individually. The alignment algorithm solves
the following problem:
\begin{equation}
\hat{\tau}_i = \arg\min_{\xx, \ee, \tau_i} \|\ee\|_1\quad \mbox{subj. to}\quad \tilde\bb\circ\tau_i = A_i\xx + \ee,
\label{eq:l1min_alignment}
\end{equation}
where $\tau_i\in T$ is in a finite-dimensional group $T$ of transformations
(e.g., affine or homography) acting on the image domain.  The sequence
$\tau_i$ converges to a transformation that aligns the test image $\tilde\bb$
with the training images from the $i$-th class $A_i$. In other words, the algorithm 
extends Lucas-Kanade iterative alignment \cite{LucasB1981} to use the
$\ell_1$-norm as a robust error function, while simultaneously estimating
the coefficients of the linear illumination model $A_i\x$.

The $ell_1$ solvers in the original system prototype were based 
on standard interior-point methods \cite{WagnerA2009-CVPR}. Their speed has been improved by ALM in a more recent paper \cite{WagnerA2011-PAMI}. However, due the high per-class cost
of the alignment step, the recognition system still fails to achieve real-time
recognition against datasets of hundreds or thousands of subjects.  In this paper,
in addition to accelerating the generic
$\ell_1$-min objectives \eqref{eq:l1min} and \eqref{eq:l1min_denoise}, we 
also discuss how to efficiently accelerate the iterative face alignment step
\eqref{eq:l1min_alignment} on multi-core CPUs and GPUs.

@@

Finally, we present extensive benchmark to compare the performance of the
generic ALM $\ell_1$-min algorithm and the corresponding face recognition
routines on massively parallel CPUs and GPUs.  To this end, a face
recognition system for security and access control applications has been implemented, 
and is diagrammed in Figure \ref{fig:pipeline}. This pipeline follows the work of \cite{WagnerA2011-PAMI}, but
the fast implementations presented in this paper are novel.
Since the parallelization of any algorithm
is highly dependent on the target hardware, we will give a brief overview of
the architecture of our test system in detail in Section
\ref{sec:concurrency}.
Finally, note that 
while this paper focuses on the parallelism available in a single (shared memory) machine,
applications targeting thousands of subjects or more, will likely require computation be
distributed over a cluster or network of computers, and this is not addressed here.
The source code of our implementations, which target NVIDIA Fermi GPUs
as well as Intel x86-64 CPUs, will be made available for peer evaluation.
\begin{figure}
\centering
{\tiny \input{figures/pipeline_simplified.tex}}
\caption{The face recognition pipeline.\vspace{-0.1in}}
\label{fig:pipeline}
\end{figure}

\subsection{Literature Review} \vspace{-0.05in}
Traditionally, $\ell_1$-min (a.k.a.
basis pursuit (BP)) has been formulated as a linear program
\cite{ChenS2001-SIAM}. Several variations of the solution are also well known
in optimization, including a noisy approximation via quadratic programming
called the LASSO \cite{TibshiraniR1996} and truncated Newton interior-point
method (TNIPM) \cite{KimS2007}.

@@

\section{Augmented Lagrangian Method}
\label{sec:ALM}
\vspace{-0.05in}
In this section, we briefly describe the ALM algorithm for $\ell_1$-min
\eqref{eq:l1min} \cite{YangA2010-ICIP} and analyze its complexity. Lagrange
multipliers have been frequently used in convex programming to eliminate
equality constraints via adding a penalty term to the cost function for
infeasible points. ALM methods differ from other penalty-based approaches by
simultaneously estimating the optimal solution and Lagrange multipliers in an
iterative fashion.  For $\ell_1$-min \eqref{eq:l1min}, the augmented Lagrange
function is defined as: 
\begin{equation} L_\mu(\xx,\yy) = \|\xx\|_1 +
\yy^T(\bb - A\xx) + \frac{\mu}{2} \| \bb-A\xx \|_2^2, 
\end{equation}
where $\mu > 0$ is a constant that penalizes infeasibility and $\yy$ is a
vector of Lagrange multipliers.

@@

 \begin{algorithm}[h]
\caption{Augmented Lagrangian Method (ALM)}
\small
{\bf INPUT:} $\bb \in \Re^m$, $A=[A_1,\cdots, A_K] \in \Re^{m \times n}$, $\tau\leftarrow \max\mbox{eig}(A^TA)$, and constant $\rho>1$.
\begin{algorithmic}[1]
\WHILE{not converged ($k = 1,2,\ldots$)} 
\STATE $t_1 \leftarrow 1$, $\zz_1 \leftarrow \xx_k$, $\uu_1 \leftarrow \xx_k$ 
\WHILE{not converged ($l = 1,2,\ldots$)} 
\STATE $\uu_{l+1}  \leftarrow \mbox{shrink}(\zz_l - \frac{1}{\tau}A^T(A\zz_l - \bb - \frac{1}{\mu_k}\yy_k), \frac{1}{\mu_k\tau})$
\STATE $t_{l+1} \leftarrow \frac{1}{2}( 1 + \sqrt{1+4t_l^2})$
\STATE $\zz_{l+1} \leftarrow \uu_{l+1}+ \frac{t_l - 1}{t_{l+1}}(\uu_{l+1} - \uu_l)$ 
\ENDWHILE 
\STATE $\xx_{k+1} \leftarrow \uu_{l+1}$ 
\STATE $\yy_{k+1} \leftarrow \yy_k + \mu_k (\bb - A\xx_{k+1})$ 
\STATE $\mu_{k+1} \leftarrow \rho\cdot\mu_k$ 
\ENDWHILE 
\end{algorithmic}

@@

\vspace{-0.05in}
\section{Hardware Concurrency} \label{sec:concurrency}
\vspace{-0.05in}
In this section we discuss the levels of concurrency available in the hardware
architectures considered in this paper, as well as other aspects of the
hardware that are important for performance.  In particular, since caches
(regions of on-chip memory) are often orders of magnitude faster than off-chip
memory, their sizes and speeds have a dramatic effect on performance.  We give a
brief overview of the caches that are available in our target architectures,
and defer discussion of their performance effects to Sections
\ref{sec:alignment_implementation_cpu} and
\ref{sec:alignment_implementation_gpu}.

Our discussion and experiments will address the most common hardware
configuration for engineering workstations: a motherboard with two quad-core
processors, and a PCI card with a single high-end GPU.  
Recognition involving more than a few hundred subjects with contemporary hardware
would require a server (or cluster) configuration with an expandable 
number of processors.  We will not address efficient parallelization for 
these systems, which may have additional challenges associated with
their non-shared memory model.
%\footnote{Non-shared memory configurations include CPU blade servers, multi-GPU servers, and clusters}.
Unless otherwise specified, all implementations utilize single precision
floating point datatypes.  
% WORKING HERE

\subsection{CPU Hardware Concurrency}
\vspace{-0.05in}
\label{sec:CPU-concurrency}
The main defining characteristics of contemporary multi-core CPU architectures
are that they have two levels of concurrency, relatively large amounts of
cache, and relatively high clock speeds. The baseline architecture for our experiments 
is a Linux workstation with two
quad-core Intel Nehalem E5530 processors clocked at 2.4 GHz.  Each processor
has its own memory interface, and is directly interfaced to half of the RAM
installed in the machine.  The amount of RAM exceeds the amount used
by the algorithms, and is not an important performance consideration.

\begin{figure}
\centering
\subfigure[The larger algorithm data structures]
{
	\includegraphics[scale=1.2]{figures/arrays}
	\label{fig:arrays}
}
\subfigure[The caches on a E5530 CPU]
{
	\includegraphics[scale=1.2]{figures/cpu_caches}
	\label{fig:cpu_caches}
}
\subfigure[The caches on a GTX480 GPU]
{
	\includegraphics[scale=1.2]{figures/gpu_caches}
	\label{fig:gpu_caches}
}
\label{fig:caches}
\caption{A visual comparison of the algorithm working set to the CPU and GPU
caches.  Note: Although the aspect ratio could not be preserved, all arrays and
caches are drawn so that the area is proportional to the size of the data. } \vspace{-0.1in}
\end{figure}

@@

For floating-point instructions, each core also has a vector processing unit
(SSE) capable of performing the same arithmetic operation on four single-precision 
floating point values simultaneously.
There are thus two important levels of concurrency that need to be exploited to
efficiently use a modern CPU: {\em core-level} concurrency and {\em SSE-level}
concurrency.

\subsection{GPU Hardware Concurrency} \vspace{-0.05in}
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
 cache, and relatively low clock speeds.  Whereas most of the transistors on
a typical CPU are dedicated either to cache or to hardware that enables higher
clock speeds (such as branch prediction, out-of-order instruction execution, etc...),
most of the transistors on a GPU are dedicated to arithmetic logic units.
For our GPU implementations we target NVIDIA Fermi GPU architecture (\eg, the GTX 480
GPU used in this paper).  An explanation of the GPU programming model (CUDA) at a useful level of completeness would take more space than is available here, so we will instead frame our discussion in terms of 
hardware capabilities. 
%\footnote{A discussion of the CUDA programming model can be found in the NVIDIA C for CUDA Programming Guide}
CUDA programmable GPUs are comprised of several
\emph{streaming multiprocessors} (SMP), each of which is roughly analogous to a CPU core.
For Fermi architecture GPU's, there are up to 16 SMP's, and each SMP is capable of executing up to 64 single precision 
floating point operations
concurrently. \footnote{In CUDA terms, the warp width is 32, and the floating pipeline can
issue up to two warps simultaneously}

@@

While the programming model for
the SIMD units of each SMP is somewhat more flexible than on the CPU,
leveraging the flexibility typically comes at the cost of reduced concurrency.
\footnote{In CUDA terms, code branches that cause warp divergence usually
result in serialization of the different groups of threads}  Therefore, for
the purposes of comparing hardware architectures, the SIMD units on the GPU are
roughly analogous to the CPU SIMD units. Thus, in summary, the GPU hardware 
provides two levels of concurrency: {\em SMP-level}
concurrency and {\em thread-level} concurrency.  Note that the GPU provides more
concurrency than a CPU at both levels (14 SMP's vs. 8 cores) and (64 wide vs. 4 wide SIMD
units).

\section{Parallelism in the Face Alignment Stage}\vspace{-0.05in}
\label{sec:alignment}

@@

Face alignment \eqref{eq:l1min_alignment} estimates an image transformation
$\tau$ that rectifies the query image $\tilde{\bb}$ with possible pose
variation w.r.t. each training class $A_i$, which leads to the minimal sparsity
in error $\ee$ after the alignment. Note that directly solving
\eqref{eq:l1min_alignment} is impossible since it is a non-convex problem
and may exhibit local minima.
However, given a good initial guess of the
transformation $\tau$ (\eg, provided by an accurate face detector), the optimal
solution for $\tau$ can be sought iteratively by linearization at each $j$th
step:
\begin{equation}
\min_{\xx, \ee, \Delta \tau_j}\|\ee\|_1\quad \subj\quad \tilde\bb\circ\tau_j +  J_j\Delta \tau_j = A_i\xx + \ee,
\end{equation}
where $J_j = \nabla_{\tau_j}(\tilde\bb\circ\tau_j)$ is the Jacobian, and
$\Delta \tau$ is a step update to $\tau$. Denote $\bb_j =
\tilde\bb\circ\tau_j$, $B_j = [A_i, -J_j]$ and $\ww^T = [\xx^T, \Delta
\tau^T]$, then the update $\Delta \tau$ can be computed by solving the
following linear program:
\begin{equation}
\min_{\ww, \ee}\|\ee\|_1\quad \subj\quad \bb_j = B_j\ww + \ee.
\label{eq:alignment-linearization}
\end{equation}
%If the warping $\tau$ is constrained to be a similarity transform,
%it can be paramaterized in such a way that the computation of $\frac{\partial \bb}{\partial \tau_p}$
%reduces to several linear vector operations involving pixel coordinates $(u,v)$, $\tau \in \Re^4$,
%and the image gradients $\f_x, \f_y \in \Re^M$, the latter of which only has to be computed
%once per test image.  
The per-class alignment algorithm via ALM is summarized in Algorithm
\ref{alg:iterative_alignment}. 
%In terms of computational complexity, one significant
%term is the computation of the matrix pseudo-inverse $B_j^\dagger$, whose
%computational cost is bounded by $O(n_i^2m + n_i^3)$. However, for per-class face
%alignment, since both $B_j$ and $J_{j}$ are very tall matrices (\ie, $m\gg n_i$
%for all $i=1, \cdots, K$ classes), we find that the computation cost is still
%dominated by the frequently computed matrix-vector multiplications in the
%inner loop.
\begin{algorithm}[ht!]
\caption{\bf (Face Alignment via ALM)} \label{alg:iterative_alignment}
\small
{\bf Input:} $\bb$, $A_i$, $\x_0 = \mathbf{0}$, $\tau_0$, and $J_0$.
\begin{algorithmic}[1]
\WHILE{not converged ($j = 1,2,\ldots$)}
\STATE Update $\bb_j \leftarrow \frac{\bb\circ \tau_{j-1}}{\|\bb\circ \tau_{j-1}\|}$; $B_j= [A_i, -J_{j-1}]$ and corresponding $(B_j^\dagger)^T$
\STATE Initialize $\ww_0 = \mathbf 0$, $\yy_0 = \mathbf 0$
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\uu_0\leftarrow \ww_{k-1}$; $\zz_0\leftarrow \ee_{k-1}$
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\zz_l \leftarrow \mbox{shrink}\left(\bb_j - B_j\uu_{l-1} + \frac{\yy_{k}}{\mu_{k-1}}, \frac{1}{\mu_{k-1}}\right)$
\STATE $\uu_l \leftarrow B_j^\dagger \left(\bb_j - \zz_{l} + \frac{\yy_{k-1}}{\mu_{k-1}} \right) $
\ENDWHILE
\STATE $\ww_k \leftarrow \uu_l$; $\ee_k \leftarrow \zz_l$
\STATE $\yy_{k} \leftarrow \yy_{k-1} + \mu_{k-1} (\bb_j - B_j\ww_{k} - \e_{k})$
\STATE $\mu_{k} \leftarrow \rho\mu_{k-1}$
\ENDWHILE
\STATE Update $\ee_j$, $\tau_j$, and $J_j$
\ENDWHILE
\end{algorithmic}
{\bf Output:} $\tau_i^*\leftarrow \tau_j, \e_i^*\leftarrow \ee_j$
\end{algorithm}

@@

\subsection{CPU Implementation} \vspace{-0.05in}
\label{sec:alignment_implementation_cpu}

@@

\subsection{Recognition Stage Implementation} \vspace{-0.05in}
Compared to problem-level
parallelism, the exclusively pixel-level parallelism in Algorithm \ref{alg:alm_rec} is
relatively straightforward to exploit:  
On the CPU, most of the operations map
well onto MKL BLAS calls, and operations that do not can be easily
parallelized using OpenMP and auto-vectorization.
